Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-11-05 paper.tex
minor fixes

\begin{abstract}
  Splitting and merging data is a requirement for most parallel or
  distributed processing operations. Naive algorithms, however,
  perform very poorly to split or merge 3D blocks from ultra-high
  resolution images, due to seek times. We introduce and
  analyze sequential algorithms that leverage memory buffering, to
  address this issue. Clustered reads and Clustered writes access
  image chunks only once, but they have to seek in the reconstructed
  image. Multiple reads and Multiple writes minimize seeks in the
  reconstructed image, but they access image chunks multiple
  times. Evaluation on a 3850x3025x3500 brain image shows that our
  algorithms perform similarly to the optimal configuration (slice
  merging) provided that enough memory is available, which solves our
  initial problem to a large extent. Besides, Multiple reads
  support on-the-fly compression of the merged image transparently but
  Clustered reads do not, due to their use of negative seeking. We
  conclude that splitting and merging large 3D images can be done
  efficiently without relying on complex data formats.
\end{abstract}

@@

We assume that the high-resolution image is split into chunks
representing 3D blocks or 2D slices that fit in memory. A dataset such
as Big~Brain would perhaps be split into 125 chunks of 600~MB. The
decision to split an image into slices or blocks, and the size of the
chunks, is up to the application. Some applications, for instance spatial filtering, would
commonly require blocks while other
ones, such as acquisition artifact removal, would rather work on
slices. Applications that process voxels individually, for instance
histogram computation or k-means clustering, could work on either
slices or blocks. Flexibility is thus required in the splitting
scheme.

We also assume that the byte organization in image files is arbitrary
but known to the algorithm. Some formats, for instance the file format
defined by the Neuroimaging Informatics Technology Initiative (NIfTI
-- \url{https://nifti.nimh.nih.gov}) store the complete image in
column-major order, that is, for an image of dimensions (i, j, k), i
is the fastest changing dimension, followed by dimension j, followed
by dimension k. Other formats, for instance HDF5-based MINC
2.0~\cite{vincent2016minc}, provide more flexibility by allowing data
to be partitioned in limited-sized chunks, each chunk being stored in
a specific order. Byte organization is obviously a critical factor of
seek time. The main idea of our algorithms will be to rearrange
ordering in memory before or after I/O operations.

The literature on this problem is remarkably scarce. Parallel and
distributed image processing has obviously been extensively studied
and used in various
platforms~\cite{miguet1991elastic,tang2007eman2,yang2008parallel,braunl2013parallel,moise2013terabyte,bajcsy2013terabyte},
but methods have focused on geometrical approaches to partition images,
and on load-balancing or task scheduling techniques. Instead, we
aim at algorithms to efficiently split or merge images regardless of
the geometry of the chunks. Although seek times are often
identified as an issue, the preferred solution is usually to optimize
data storage formats for a certain application.  For instance, the
Open Connectome Data Cluster~\cite{burns2013open} is a data warehouse
system that allows users to retrieve specific 3D blocks from large
image datasets. It reduces seek times through a specific file format
based on space-filling curves, which elegantly preserves spatial
proximity on disk. On the contrary, we are searching for algorithms
that would reduce the seek time regardless of the data format, so that
applications with flexible splitting schemes can be served

@@

The buffer might be a slightly complex data structure, for instance an
associative array or a Python dictionary, capable of storing multiple
disjoint sequences of contiguous bytes without having to allocate
memory for the bytes between such
sequences. Figure~\ref{fig:cluster-reads-buffer} illustrates how the
buffer would fill up for the two first blocks in a reconstructed
image, assuming that blocks are of size 4x4x4.

@@

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case1-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case2-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case3.pdf_tex}
\caption{Memory-load configurations in Clustered reads, leading to
  different number of seeks. Red blocks need seeking before each of
  their rows ($d^2$ seeks). Blue blocks need seeking before each of
  their slices ($d$ seeks). Green blocks need only a single
  seek. Grey, dashed, transparent blocks represent the contiguous
  memory loads and are added for the sake of visualization.}
\label{fig:cluster-reads}
\end{figure}
Clustered reads focus on the three memory load
configurations represented in Figure~\ref{fig:cluster-reads}, that is,
the amount of memory $m'$ used by the algorithm is rounded down to the
closest number of complete blocks (Case 1), of complete block rows (Case
2) or of complete block slices (Case 3). This is in general reasonable
since adding an incomplete row to a set of complete ones multiplies
the number of required seeks by $d$, as illustrated in
Figure~\ref{fig:avoided-configurations}-Left. In some cases though,
rounding $m$ down to $m'$ might increase the number of required
memory loads to a point that the overall number of seeks also
increases. Such cases are, however, slightly unusual and their
complete description requires extensive calculations involving modulo
arithmetic, which we felt were unwieldy to report here.
\begin{figure}
  \centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/incomplete-rows.pdf_tex}
\quad \quad \quad
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/overlap.pdf_tex}
\caption{Configurations that increase the number of seeks per
  memory-load and are thus deliberately avoided by Clustered
  reads. Left: configuration with incomplete block rows (multiplies
  the number of seeks by $d$ compared to complete block rows). Right:
  configuration with block rows that overlap multiple block slices
  (multiplies the number of seeks by $2$ compared to non-overlapping
  configurations).}
\label{fig:avoided-configurations}
\end{figure}

@@

We implemented lossless compression for all algorithms using Python's
gzip library. Compression is done on the fly, that is, while the data
is being read or written. On-the-fly compression of large datasets is
a challenge when extensive seeking is involved as the gzip library has
to decompress all the data until the seek position to read from it,
making access time a linear function of the seek position. This is
potentially an issue for Clustered writes, which could be addressed by
indexing techniques such as described in~\cite{rajna2015speeding} for
the NiFTI format and implemented in
\url{https://github.com/pauldmccarthy/indexed\_gzip}. Multiple writes
are not impacted since they do not seek in the large image.

@@

\subsection{Data}
We used the 3850x3025x3500 Big~Brain image split in 125
non-overlapping chunks of size 770x605x700 with 2 bytes per voxel
(total size uncompressed is 75.92~GB). Big~Brain is a reference brain
based on the reconstruction of 7404 histological sections at nearly
cellular resolution of 20 micrometers~\cite{amunts2013bigbrain}. It is
a freely, publicly available tool with numerous applications in
neurosciences and neurosurgery.

We used the blocks of the 2015 Big~Brain release with 40-micrometer
isotropic resolution available at
\url{ftp://bigbrain.loris.ca/BigBrainRelease.2015/3D\_Blocks/40um}. We
converted them to NiFTI 1.0 using Nibabel and left them uncompressed.
We generated NiFTI slabs of size 3850x3025x28 from the reconstructed
image, using our naive split algorithm.

@@

Compressed blocks and slices were also merged into a compressed
image. On-the-fly gzip compression was used for Naive slices, Buffered
slices and Multiple reads. As explained before, Clustered reads could
only be applied to compressed data while in Case 3, i.e., for 16~GB,
and Naive blocks could not be used at all with on-the-fly
compression. To give a reference, with used Naive blocks with
\emph{offline} compression, that it, we wrote an uncompressed image
and compressed it sequentially afterwards.

@@

>>>>

\begin{abstract}
  Splitting and merging data is a requirement for most parallel or
  distributed processing operations. Naive algorithms to split and
  merge 3D blocks from ultra-high resolution images perform very poorly,
  as a result of seek times. In contrast, naive algorithms to split and 
  merge 2D slices or 3D slabs perform optimally as seek time is significantly minimized.
  We introduce and analyze sequential algorithms (Clustered reads, Multiple reads, Clustered write and Multiple writes)
  that leverage memory buffering to
  address this issue. Clustered reads and Clustered writes, access
  image chunks only once, but they have to seek in the reconstructed
  image. Multiple reads and Multiple writes minimize seeks in the
  reconstructed image, but they access image chunks multiple
  times. Evaluation on a 3850x3025x3500 brain image shows that our
  algorithms perform similarly to the optimal configuration provided that enough memory is available. 
  Additionally, Multiple reads support on-the-fly compression of the 
  merged image transparently but Clustered reads do not, due to their use of negative seeking. We
  conclude that splitting and merging large 3D images can be done
  efficiently without relying on complex data formats.
\end{abstract}

@@

We assume that the high-resolution image is split into chunks
representing 3D blocks or 2D slices that fit in memory. A dataset such
as Big~Brain would perhaps be split into 125 chunks of 600~MB. The
decision to split an image into slices or blocks, and the size of the
chunks, is up to the application. For instance, spatial filtering would
commonly require blocks, whereas slices would be preferable for acquisition artifact removal. 
Applications that process voxels individually, for instance
histogram computation or k-means clustering, could work on either
slices or blocks. Flexibility is thus required in the splitting
scheme.

We also assume that the byte organization in image files is arbitrary
but known to the algorithm. Some formats, for instance the file format
defined by the Neuroimaging Informatics Technology Initiative (NIfTI
-- \url{https://nifti.nimh.nih.gov}) store the complete image in
column-major order, that is, elements belonging to the same column are
stored in a contiguous order. Other formats, such as HDF5-based MINC
2.0~\cite{vincent2016minc}, provide more flexibility by allowing data
to be partitioned in limited-sized chunks, each chunk being stored in
a specific order. Byte organization is obviously a critical factor of
seek time. The main idea of our algorithms will be to rearrange
ordering in memory before or after I/O operations.

The literature on this problem is remarkably scarce. Parallel and
distributed image processing has obviously been extensively studied
and used in various
platforms~\cite{miguet1991elastic,tang2007eman2,yang2008parallel,braunl2013parallel,moise2013terabyte,bajcsy2013terabyte},
but methods have focused on geometrical approaches to partition images,
and on load-balancing or task scheduling techniques. Instead, we
aim at algorithms to efficiently split or merge images regardless of
the geometry of the chunks. Although seek times are often
identified as an issue, the preferred solution is usually to optimize
data storage formats for a certain application.  For instance, the
Open Connectome Data Cluster~\cite{burns2013open} is a data warehouse
system that allows users to retrieve specific 3D blocks from large
image datasets. It reduces seek times through a specific file format
based on space-filling curves, which elegantly preserves spatial
proximity on disk. On the contrary, we are searching for algorithms
that would reduce the seek time regardless of the data format, such that
applications with flexible splitting schemes can be served.

@@

The buffer, capable of storing multiple
disjoint sequences of contiguous bytes without having to allocate
memory for the bytes between such
sequences, can be represented by an
associative array or a Python dictionary.
Figure~\ref{fig:cluster-reads-buffer} illustrates how the
buffer would fill up for the two first blocks in a reconstructed
image, assuming, for the sake of this particular example, that blocks are of size 4x4x4.

@@

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case1-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case2-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case3.pdf_tex}
\caption{Memory-load configurations in Clustered reads, leading to
  different number of seeks. Red blocks need seeking before each of
  their rows ($d^2$ seeks). Blue blocks need seeking before each of
  their slices ($d$ seeks). Green blocks need only a single
  seek. Grey, dashed, transparent blocks represent the contiguous
  memory loads and are added for the sake of visualization.}
\label{fig:cluster-reads}
\end{figure}
Clustered reads focus on the three memory load
configurations represented in Figure~\ref{fig:cluster-reads}:
the amount of memory $m'$ used by the algorithm is rounded down to the
closest number of complete blocks (Case 1), of complete block rows (Case
2) or of complete block slices (Case 3). This is in general reasonable
since adding an incomplete row to a set of complete ones multiplies
the number of required seeks by $d$, as illustrated in
Figure~\ref{fig:avoided-configurations}-Left. In some cases though,
rounding $m$ down to $m'$ might increase the number of required
memory loads to a point that the overall number of seeks also
increases. Such cases are, however, slightly unusual and their
complete description requires extensive calculations involving modulo
arithmetic, which we felt were unwieldy to report here.
\begin{figure}
  \centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/incomplete-rows.pdf_tex}
\quad \quad \quad
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/overlap.pdf_tex}
\caption{Configurations that increase the number of seeks per
  memory-load and are thus deliberately avoided by Clustered
  reads. Left: configuration with incomplete block rows (multiplies
  the number of seeks by $d$ compared to complete block rows). Right:
  configuration with block rows that overlap multiple block slices
  (multiplies the number of seeks by $2$ compared to non-overlapping
  configurations).}
\label{fig:avoided-configurations}
\end{figure}

@@

We implemented lossless compression for all algorithms using Python's
gzip library. Compression is done on the fly, that is, while the data
is being read or written during splitting and merging. 
On-the-fly compression of large datasets is
a challenge when extensive seeking is involved as the gzip library has
to decompress all the data until the seek position to read from it,
making access time a linear function of the seek position. This is
potentially an issue for Clustered writes, which could be addressed by
indexing techniques such as described in~\cite{rajna2015speeding} for
the NiFTI format and implemented in
\url{https://github.com/pauldmccarthy/indexed\_gzip}. Multiple writes
are not impacted since they do not seek in the large image.

@@

\subsection{Data}
We used the 3850x3025x3500 Big~Brain image split into 125
non-overlapping chunks of size 770x605x700, with 2 bytes per voxel
(total size uncompressed is 75.92~GB). The Big~Brain image was also split
into 125 non-overlapping slabs of size 3850x3025x28 for our experiments.

Big~Brain is a reference brain
based on the reconstruction of 7404 histological sections at nearly
cellular resolution of 20 micrometers~\cite{amunts2013bigbrain}. It is
a freely, publicly available tool with numerous applications in
neurosciences and neurosurgery.

We used the blocks of the 2015 Big~Brain release with 40-micrometer
isotropic resolution available at
\url{ftp://bigbrain.loris.ca/BigBrainRelease.2015/3D\_Blocks/40um}. We
converted them to NiFTI 1.0 using Nibabel and left them uncompressed.
These blocks were then used to reconstruct the Big~Brain that was split into our desired block and slab
configurations using our naive splitting algorithms.

@@

Compressed blocks and slabs were also merged into a compressed
image. On-the-fly gzip compression was used for Naive slices, Buffered
slices and Multiple reads. As explained before, Clustered reads could
only be applied to compressed data while in Case 3, i.e., for 16~GB,
and Naive blocks could not be used at all with on-the-fly
compression. To give a reference, we used Naive blocks with
\emph{offline} compression, that is, we wrote an uncompressed image
and compressed it sequentially afterwards.

@@

It should be noted that in some cases, splitting and merging can be considered to be an independent process
from the other tasks. As such, the performance impact splitting and 
merging has on the main tasks depends largely on the duration of the main tasks.
